---
title: "Predict detection probability at given distance"
author:
  - name: Eric Rexstad 
    affiliation: CREEM, Univ of St Andrews
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_depth: 2
    base_format: rmarkdown::html_vignette
pkgdown:
  as_is: true 
bibliography: detfct.bib
csl: ../apa.csl     
vignette: >
  %\VignetteIndexEntry{Predict detection probability at given distance}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Motivation

Having fitted a detection function model to distance sampling data, often there is interest in using the fitted model to predict detection probability for combinations of predictors (including distance). 

Commonly this is done when covariates other that distance are included in the model. For example, researchers may want to know the difference in predicted detection probability at a given distance among different habitat types. It is this scenario that lead to the development of this case study. Credit to Ward Langeraert for bringing this to our attention.

We construct a simple simulated line transect distance sampling survey, sampling a population consisting of four types: two sexes and two ages. Each of the four categories of animals have distinct simulated detection functions. We generate a single survey of this population and fit a detection function containing an `agesex` covariate, consistent with how the simulated data were generated.

The main focus of this case study is interrogating the fitted model to a) visualise the fitted detection function model for each of the four age/sex classes and b) focus upon the differences in predicted detection probability between age/sex classes at a specified perpendicular distance. However, the application of the approach demonstrated in this case study is not limited only to those situations.

# Simulate a data set

The simulation of a data set is accomplished using the `dsims` package @dsimspkg. Details of the [`dsims`](https://distancesampling.org/dsims/index.html) package will not be described here, interested readers should follow the previous link.

```{r, fig.width=7, fig.asp=.75, fig.cap="Simulated detection functions for four age x sex classes."}
library(dsims)
set.seed(19511)
study.region <- make.region()
density <- make.density(region = study.region)
covariate.list <- list()
# The population equal proportion in four age/sex classes
covariate.list$agesex <- list(data.frame(level = c("youngmale", "youngfemale", "adultmale", "adultfemale"), 
                                          prob = rep(0.25,4)))
pop.desc <- make.population.description(region = study.region, 
                                        density = density, 
                                        N = 400,
                                        covariates = covariate.list,
                                        fixed.N = TRUE)
cov.params <- list()
truncation_distance <- 50
cov.params$agesex = data.frame(level = c("youngmale", "youngfemale", "adultmale", "adultfemale"), 
                            param = c(0.2, 0.7, 1.0, 1.3))
detect.cov <- make.detectability(key.function = "hn" ,
                                 scale.param = 10,
                                 cov.param = cov.params, 
                                 truncation = truncation_distance)
plot(detect.cov, pop.desc, lwd=3)
design <- make.design(region = study.region,
                      transect.type = "line",
                      design = "systematic",
                      spacing = 100,
                      truncation = truncation_distance)
analysis.cov <- make.ds.analysis(dfmodel = list(~agesex),
                                 key = "hn",
                                 truncation = truncation_distance)
my.simulation <- make.simulation(reps = 999, 
                                 design = design,
                                 population.description = pop.desc,
                                 detectability = detect.cov,
                                 ds.analysis = analysis.cov)
my.survey <- run.survey(my.simulation)
ds_model <- analyse.data(analysis.cov, my.survey)
```

The plot created in the above chunk shows the simulated detection functions for each age/sex class in the simulated population. The last object created in the above chunk is of type `dsmodel`, equivalent to objects created by the `ds()` function in the `Distance` package. It is this object (`ds_model`) that you would have created in your analysis and want to use to predict detection probabilities from the fitted model.

# Interrogate `ds_model$ddf`

To produce our predicted detection probabilities, we need merely one component of the `ds_model` object. It is called `ddfobj` and it is located deep within `ds_model`, specificially `ds_model$ddf$ds$aux$ddfobj`

The function that does the work of producing estimated detection probabilities from a `ddfobj` along with a detection distance and an index to the set of predictor covariates is called `detfct` in the `mrds` package.

Code below isolated the `ddfobj` within `ds_model` just to make life easier. Then the index of the first observation of each age/sex combination is found within the design matrix (because the design matrix is formed after any truncation takes place in model fitting).

```{r}
ddfobj <- ds_model$ddf$ds$aux$ddfobj
ym.indx <- which(ddfobj$xmat$agesex=="youngmale")[1]
yf.indx <- which(ddfobj$xmat$agesex=="youngfemale")[1]
am.indx <- which(ddfobj$xmat$agesex=="adultmale")[1]
af.indx <- which(ddfobj$xmat$agesex=="adultfemale")[1]
```

This completes the preliminaries to use `detfct` for predicting detection probabilities.

## Predict detection probability at a specific distance

The next code chunk uses `detfct` to predict detection probabilities for each age/sex class at a single perpendicular distance.

```{r}
mydist <- 35
Young_male <- mrds::detfct(mydist, ddfobj, index = ym.indx)
Young_female <- mrds::detfct(mydist, ddfobj, index = yf.indx)
Adult_male <- mrds::detfct(mydist, ddfobj, index = am.indx)
Adult_female <- mrds::detfct(mydist, ddfobj, index = af.indx)
outtable <- data.frame(Age_sex_class=c("Young male", "Young female", "Adult male", "Adult female"),
                       Predicted_p=c(Young_male, Young_female, Adult_male, Adult_female))
knitr::kable(outtable, digits = 3, 
             caption="Predicted detection probabilities at 35m")
```

# Compute detection probabilities at range of distances

We will create a code chunk that duplicates the capability of the [`add_df_covar_line` function](https://distancesampling.org/mrds/reference/add.df.covar.line.html) found in the `mrds` package. The code plots a detection function for each of the four age/sex classes over a range of requested distances using the indices calculated in the previous chunk.

```{r}
prediction_distances <- seq(0, 50) 
g_x_ym <- mrds::detfct(prediction_distances, ddfobj, index = ym.indx)
g_x_yf <- mrds::detfct(prediction_distances, ddfobj, index = yf.indx)
g_x_am <- mrds::detfct(prediction_distances, ddfobj, index = am.indx)
g_x_af <- mrds::detfct(prediction_distances, ddfobj, index = af.indx)
```

The vectors of predicted detection probabilities are plotted in the following chunk. The last section of the chunk adds the calculated probabilities at the 35m distance to the plot.

```{r, fig.width=7, fig.asp=.75, fig.cap="Predicted detection probabilities for four age x sex classes; with those at 35m highlighted."}
plot(ds_model, pl.col="white", cex=0.7)
lines(prediction_distances, g_x_ym, col="darkgreen", lwd=2)
lines(prediction_distances, g_x_yf, col = "green", lwd=2)
lines(prediction_distances, g_x_am, col = "darkblue", lwd=2)
lines(prediction_distances, g_x_af, col = "brown", lwd=2)
legend("topright", title="Age/sex class", lwd=2,
       legend=c("young male", "young female", "adult male", "adult female"),
       col=c("darkgreen", "green", "darkblue", "brown"))
#----------
abline(v=mydist, lwd=3)
segments(0, Young_male, mydist, Young_male, col="darkgreen")
segments(0, Young_female, mydist, Young_female, col="green")
segments(0, Adult_male, mydist, Adult_male, col="darkblue")
segments(0, Adult_female, mydist, Adult_female, col="brown")
text(0, Young_male, round(Young_male,3), adj=c(-.1, -.3), cex=0.7)
text(0, Young_female, round(Young_female,3), adj=c(-.1, -.3), cex=0.7)
text(0, Adult_male, round(Adult_male,3), adj=c(-.1, -.3), cex=0.7)
text(0, Adult_female, round(Adult_female,3), adj=c(-.1, -.3), cex=0.7)
```

